---
title: "TMA4300: Exercise 1"
author: "Jim Totland, Martin Tufte"
date: "1/29/2022"
output: pdf_document
fig_width: 3
fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 4, fig.align = "center")
library(tidyverse)
```

# Problem A
## A.1
The exponential distribution has cumulative density function (CDF)  

$$
F(x) = 1 - e^{-\lambda x},
$$

with rate parameter $\lambda$. By defining $u := F(x)$, we can express $x$ as 

$$
x = -\frac{1}{\lambda}\ln(1-u) =: F^{-1}(u).
$$
This means that we can use the \textit{inversion method} to simulate from the exponential distribution. I.e., we let $U \sim \mathcal{U}_{[0,1]}$ and calculate $X = F^{-1}(U)$. Then, $X \sim \mathrm{Exp}(\lambda)$. The function which simulates the exponential distribution is given below.

```{r}
sim.exp <- function(rate, n){
  u <- runif(n,0,1)
  return(-1/rate * log(1 - u))
}
```
Next, we need to check if this gives reasonable results by comparing our simulated values to the theoretical knowledge.
```{r, cache=TRUE}
rate <- 0.5
n <- 1000
sim <- data.frame(x = sim.exp(rate, n))
x = seq(from = 0, to = 15, by = 0.1)
exact <- data.frame(x = x, y = rate*exp(-rate*x))

ggplot(sim) + 
  geom_histogram(aes(x = x, y = ..density..),
                 alpha = 0.8, fill = "#69b3a2", binwidth = 0.1) + 
  geom_line(data = exact , aes(x = x, y = y), color = "#e0474c")  +
  geom_label(
    label="Simulated", 
    x=14,
    y=0.5,
    label.padding = unit(0.55, "lines"), # Rectangle size around label
    label.size = 0.35,
    color = "white",
    fill= "#69b3a2"
  ) +
  geom_label(
    label="Exact", 
    x=14,
    y=0.42,
    label.padding = unit(0.55, "lines"), # Rectangle size around label
    label.size = 0.35,
    color = "white",
    fill = "#e0474c"
  ) +
  xlim(0,17)
```
We also compare the estimated mean and variance to first and second central moments of the exponential distribution

```{r}
# mean
data.frame(n.samples = 1:n,
           mean = cumsum(sim$x)/(1:n)) %>%
  ggplot() + geom_line(aes(n.samples, mean)) +
  geom_hline(yintercept = 1/rate, color = "red") +
  ggtitle("mean")
  
# variance
data.frame(n.samples = 1:n,
           mean = cumsum(sim$x)/(1:n),
           mean2 = cumsum(sim$x^2)/(1:n)) %>%
  mutate(var = mean2-mean^2) %>%
  ggplot() + geom_line(aes(n.samples, var)) +
  geom_hline(yintercept = 1/rate^2, color = "red") +
  ggtitle("variance")
```
We observe that the the sampled mean and variance approach the theoretical values as the number of samples grows larger. 

## A.3
### a)
To find the value of $c$, we integrate the density over the entire domain and equate the result to 1:
$$
1 = \int_{-\infty}^{\infty}f(x)dx = c\int_{-\infty}^{\infty}\frac{e^{\alpha x}}{(1 + e^{\alpha x})^2}dx.
$$
To progress from here, we introduce the substitution, $v = e^{\alpha x}$, which gives

$$
1 = \frac{c}{\alpha} \int_0^{\infty} \frac{dv}{(1 + v)^2} = \frac{c}{\alpha}(-\frac{1}{1 + v})\Big|_0^\infty =\frac{c}{\alpha}.
$$
Consequently, $c = \alpha$.

### b)
The CDF is defined as follows,
$$
\begin{split}
F(x) &= \int_{-\infty}^xf(z)dz =  \int_0^{\exp(\alpha x)} \frac{dv}{(1 + v)^2} \\
 &= 1-\frac{1}{1 + \exp(\alpha x)} = \frac{\exp(\alpha x)}{1 + \exp(\alpha x)},
\end{split}
$$

where we have used the same substitution as earlier, namely $v = e^{\alpha z}$. We notice that $F(x)$ is the Sigmoid function, which has the well known logit-function as its inverse. I.e.,
$$
F^{-1}(x) = \frac{1}{\alpha}\ln\left(\frac{x}{1-x}\right).
$$

### c)

Since we have an analytic expression for the inverse, we can again use the \textit{inversion method} to sample from $f$. The sampling-function is given below.

```{r}
sim.sigm <- function(alpha, n){
  u <- runif(n,0,1)
  return(1/alpha * log(u/(1-u)))
}
```

```{r}
alpha <- 2
n <- 1000
sim <- data.frame(x = sim.sigm(alpha, n))
x = seq(from = -10, to = 10, by = 0.1)
exact <- data.frame(x = x, y = alpha*exp(alpha*x)/(1 + exp(alpha*x))^2)

ggplot(sim) + 
  geom_histogram(aes(x = x, y = ..density..),
                 alpha = 0.8, fill = "#69b3a2", binwidth = 0.1) + 
  geom_line(data = exact , aes(x = x, y = y), color = "#e0474c")  
```

To compare the simulated values with the theoretical mean and variance, we first need to compute the expression of these moments. \textcolor{red}{Think maybe the mean is undefined??}

## A.4
Below is our implementation of the Box-Muller algorithm.
```{r}
box.mul <- function(n){
  odd <-  FALSE
  if(n %% 2 != 0){
    odd <-  TRUE
    n <- n + 1
  }
  
  x1 <- 2*pi * runif(n/2, 0, 1)
  x2 <- sim.exp(0.5, n/2)
  
  y1 <- sqrt(x2)*cos(x1)
  y2 <- sqrt(x2)*sin(x1)
  
  concat <- c(y1,y2)
  if(odd){
    return(head(concat, -1))
  }
  else{
    return(concat)
  }
}
```
As usual, we compare the results of a simulation against the theoretical distribution.
```{r}
n <- 1000
sim.box.mul <- data.frame(x = box.mul(n))
x = seq(from = -5, to = 5, by = 0.1)
exact <- data.frame(x = x, y = 1/sqrt(2*pi)*exp(-1/2*x^2))

ggplot(sim.box.mul) + 
  geom_histogram(aes(x = x, y = ..density..),
                 alpha = 0.8, fill = "#69b3a2", binwidth = 0.1) + 
  geom_line(data = exact , aes(x = x, y = y), color = "#e0474c")  
```
Finally, we compare the mean and variance resulting from the simulation with the theoretical values. 
```{r}
# mean
data.frame(n.samples = 1:n,
           mean = cumsum(sim.box.mul$x)/(1:n)) %>%
  ggplot() + geom_line(aes(n.samples, mean)) +
  geom_hline(yintercept = 0, color = "red") +
  ggtitle("mean")
  
# variance
data.frame(n.samples = 1:n,
           mean = cumsum(sim.box.mul$x)/(1:n),
           mean2 = cumsum(sim.box.mul$x^2)/(1:n)) %>%
  mutate(var = mean2-mean^2) %>%
  ggplot() + geom_line(aes(n.samples, var)) +
  geom_hline(yintercept = 1, color = "red") +
  ggtitle("variance")
```


## A.5

# Problem B
## B.1
### (a)

## B.2
### (a)
We repeat that 
$$
f^*(x) = \begin{cases} x^{\alpha-1}e^{-x}, \quad &x>0, \\
0, \quad &\text{otherwise}.
\end{cases}
$$
To find $a = \sqrt{\sup_{x\ge 0} f^*(x)}$, we first note that $f^*(x = 0) = 0$ and only consider
$$
\sup_{x > 0} f^*(x),
$$
which amounts to solving

$$
\begin{split}
\frac{d}{dx} f^*(x)= 0 \\
&\iff (\alpha-1)x^{\alpha -2}e^{-x} - x^{\alpha -1}e^{-x} = 0 \\
&\iff x=\alpha-1>0,
\end{split}
$$
which clearly is a global maximum. Consequently,
$$
a = \sqrt{\alpha - 1}.
$$

Analogously, to find  $b_+ = \sqrt{\sup_{x\ge 0} x^2f^*(x)} = \sqrt{\sup_{x\ge 0}g(x)}$, we note that $g(z = 0) = 0$ and only consider $\sup_{x > 0}g(x)$, which amounts to solving

$$
\begin{split}
\frac{d}{dx} g(x)= 0 \\
&\iff (\alpha+1)x^{\alpha}e^{-x} - x^{\alpha + 1}e^{-x} = 0 \\
&\iff x=\alpha + 1>0,
\end{split},
$$
which clearly is a global maximum for $x \ge 0$. Consequently,
$$
b_+ = \sqrt{\alpha + 1}.
$$

Finally, $b_{-} = - \sqrt{\sup_{x \le 0} x^2f^*(x)} = - \sqrt{\sup_{x \le 0} g(x)}$

### (b)

```{r}

alpha <-  1
a <-  sqrt(1-alpha)
b.minus <- 0
b.plus <- sqrt(alpha + 1)

rou.gamma <- function(n, a, b.minus, b.plus, alpha){
  count <-  0
  tries <- 0
  result <- rep(0,n)
  while(count < n){
    x1 <- a * runif(1, 0, 1)
    x2 <- b.minus + b.plus * runif(1, 0, 1) 
    
    if(log(x1) <= 0.5*((alpha - 1)*log(x2/x1) - x2/x1)){
      result[count + 1] = x2/x1
      count <- count + 1
    }
    tries = tries + 1
  }
  return(data.frame(sim = result, tries = tries))
}
```

First, we check that the simulation algorithm gives reasonable results:
```{r}
n <- 5000
alpha <-  3
a <-  sqrt(alpha - 1)
b.minus <- 0
b.plus <- sqrt(alpha + 1)

sim.gamma <- rou.gamma(n, a, b.minus, b.plus, alpha)
x = seq(from = 0.1, to = 12, by = 0.1)
exact <- data.frame(x = x, y = 1/gamma(alpha) * x^(alpha - 1) * exp(-x))

ggplot(sim.gamma) + 
  geom_histogram(aes(x = sim, y = ..density..),
                 alpha = 0.8, fill = "#69b3a2", binwidth = 0.1) + 
  geom_line(data = exact , aes(x = x, y = y), color = "#e0474c") 

```
```{r}
library(pracma)
n <- 1000
alpha.vec <- logseq(1.01, 2000, n = 500)
trial.vec <- rep(0, 100)
for(i in 1:length(alpha.vec)){
  alpha = alpha.vec[i]
  sim.gamma <- rou.gamma(n, a, b.minus, b.plus, alpha)
  trial.vec[i] = sim.gamma$tries[1]
}

plot(log(alpha.vec, base = 10), log(trial.vec, base = 10), ylim = c(2.7,4))
abline(h = 3, col="blue")
```
The number of necessary tries is highest for low values of $\alpha$, and decreases fast, before it stabilizes around $10^{3.2}$. \textcolor{red}{mer tolkning??}

## B.3
We recall that the gamma distribution has probability density function (PDF)

$$
f(x) = \frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha -1}e^{-\beta x}.
$$
We use Marsaglia and Tsang’s method to generate from the gamma distribution:
```{r}
sim.gamma <- function(n, alpha, beta){
  d <- alpha - 1/3
  c <- 1/sqrt(9*d)
  count <- 0
  result <- rep(0, n)
  while(count < n){
    z <- box.mul(1)
    u <- runif(1, 0, 1)
    v <- (1 + c*z)^3
    if(z > -1/c && log(u) < 0.5*z^2 + d - d*V + d*log(V)){
      result[count + 1] = d*v
      count <- count + 1
    }
  }
  return(1/beta * result)
}
```

Check that it works god:...

## B.4
We repeat that $X \sim \text{Gamma}(\alpha, 1)$ and  $Y \sim \text{Gamma}(\beta, 1)$, and note that their joint density is $f_{X,Y}(x,y) = f_X(x)f_Y(y)$, since they are independent. Next, we define $Z = \frac{X}{X + Y}$ and also introduce $V = X + Y$. Then, $X = ZV$ and $Y=V(1-Z)$, and, by the transformation formula, the joint density of $Z$ and $V$ is

$$
\begin{split}
f_{Z, V}(z,v) &= f_{X, Y}(zv, v(1-z)) \cdot \begin{vmatrix} v & z \\
-v & 1-z\end{vmatrix} \\
&= (\Gamma(\alpha) \Gamma(\beta))^{-1}\cdot(zv)^{\alpha-1} e^{-zv} \cdot (v(1-z))^{\beta -1}e^{-v(1-z)}\cdot[v(1-z) + vz] \\
&=(\Gamma(\alpha) \Gamma(\beta))^{-1} \cdot z^{\alpha-1}(1-z)^{\beta - 1}\cdot v^{[\alpha + \beta] - 1} \cdot e^{-v}.
\end{split}
$$
Thus, the marginal distribution of $Z$ is given as

$$
\begin{split}
\int_0^\infty f_{Z, V}(z,v)dv &= \Gamma(\alpha) \Gamma(\beta))^{-1} \cdot z^{\alpha-1}(1-z)^{\beta - 1} \int_0^\infty v^{[\alpha + \beta] - 1} \cdot e^{-v} dv \\
&= \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} z^{\alpha-1}(1-z)^{\beta - 1},
\end{split}
$$
which is what we wanted show.
