---
title: "TMA4300: Exercise 1"
author: "Jim Totland, Martin Tufte"
date: "1/29/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Problem A
## A.1
The exponential distribution has cumulative density function (CDF)  

$$
F(x) = 1 - e^{-\lambda x},
$$

with rate parameter $\lambda$. By defining $u := F(x)$, we can express $x$ as 

$$
x = -\frac{1}{\lambda}\ln(1-u) =: F^{-1}(u).
$$
This means that we can use the \textit{inversion method} to simulate from the exponential distribution. I.e., we let $U \sim \mathcal{U}_{[0,1]}$ and calculate $X = F^{-1}(U)$. Then, $X \sim \mathrm{Exp}(\lambda)$. The function which simulates the exponential distribution is given below.

```{r}
sim.exp <- function(rate, n){
  u <- runif(n,0,1)
  return(-1/rate * log(1 - u))
}
```
Next, we need to check if this gives reasonable results by comparing our simulated values to the theoretical knowledge.
```{r, cache=TRUE}
rate <- 0.5
n <- 10000
sim <- data.frame(x = sim.exp(rate, n))
x = seq(from = 0, to = 15, by = 0.1)
exact <- data.frame(x = x, y = rate*exp(-rate*x))

ggplot(sim) + 
  geom_histogram(aes(x = x, y = ..density..),
                 alpha = 0.8, fill = "#69b3a2", binwidth = 0.1) + 
  geom_line(data = exact , aes(x = x, y = y), color = "#e0474c")  +
  geom_label(
    label="Simulated", 
    x=14,
    y=0.5,
    label.padding = unit(0.55, "lines"), # Rectangle size around label
    label.size = 0.35,
    color = "white",
    fill= "#69b3a2"
  ) +
  geom_label(
    label="Exact", 
    x=14,
    y=0.42,
    label.padding = unit(0.55, "lines"), # Rectangle size around label
    label.size = 0.35,
    color = "white",
    fill = "#e0474c"
  )
```
We also compare the estimated mean and variance to first and second central moments of the exponential distribution

```{r}
# mean
data.frame(n.samples = 1:n,
           mean = cumsum(sim$x)/(1:n)) %>%
  ggplot() + geom_line(aes(n.samples, mean)) +
  geom_hline(yintercept = 1/rate, color = "red") +
  ggtitle("mean")
  
# variance
data.frame(n.samples = 1:n,
           mean = cumsum(sim$x)/(1:n),
           mean2 = cumsum(sim$x^2)/(1:n)) %>%
  mutate(var = mean2-mean^2) %>%
  ggplot() + geom_line(aes(n.samples, var)) +
  geom_hline(yintercept = 1/rate^2, color = "red") +
  ggtitle("variance")
```


